---
title: "stormdata"
author: "sstolk"
date: "25 January 2015"
output: html_document
---
**Synopsis**
##
The goal of this analysis is to identify the types of severe weather events that cause the most economic damage and the most damage to public health. For the analysis, data on crop damage, property damage, injuries and fatalities caused by types of severe weather events. This data was collected by the National Weather Service.

**Data Processing**
##
The first step was to load the target data into R. 
```{r}
stormdata <- read.csv("./repdata_data_StormData.csv")
head(stormdata)
head(unique(stormdata$EVTYPE))
```
This shows the key fields to be used in assessing harm caused by severe weather events. In accordance with the analysis described above, these can be divided into economic damage and population health damage. Crop damage and property damage will be classed as economic damage, while fatalities and injuries will be classed as population health damage. 
To create a table of the economic damage data:
```{r}
propData <- subset(stormdata, subset=stormdata$PROPDMG>0)
cropData <- subset(stormdata, subset=stormdata$CROPDMG>0)
econData <- rbind(propData, cropData)
```
To create a table of to population health damage data:
```{r}
fatalData <- subset(stormdata, subset=stormdata$FATALITIES>0)
injuryData <- subset(stormdata, subset=stormdata$INJURIES>0)
harmData <- rbind(fatalData, injuryData)
```

**Results**
##
The following is a table of results for property damage by severe weather event.
```{r}
totalProp <- aggregate(propData$PROPDMG, by=list(propData$EVTYPE), FUN=sum)
colnames(totalProp) <- c("Event", "Property_Damage")
totalProp <- totalProp[order(totalProp$Property_Damage, decreasing=TRUE),]
head(totalProp)
```
The table shows that tornados cause the most property damage, followed by flash floods and high winds associated with thunderstorms. 
```{r}
#                Event Property_Damage
#334           TORNADO       3212258.2
#51        FLASH FLOOD       1420124.6
#348         TSTM WIND       1335965.6
#64              FLOOD        899938.5
#296 THUNDERSTORM WIND        876844.2
#106              HAIL        688693.4
```
The economic damage also includes damage to crops, as shown in the following table. 
```{r}
> totalCrop <- aggregate(cropData$CROPDMG, by=list(cropData$EVTYPE), FUN=sum)
> colnames(totalCrop) <- c("Event", "Crop_Damage")
> totalCrop <- totalCrop[order(totalCrop$Crop_Damage, decreasing=TRUE),]
> head(totalCrop)
```
The most damaging weather event is hail, followed by flash floods and floods. 

```{r}
#                Event Crop_Damage
#42               HAIL   579596.28
#23        FLASH FLOOD   179200.46
#27              FLOOD   168037.88
#115         TSTM WIND   109202.60
#107           TORNADO   100018.52
#94  THUNDERSTORM WIND    66791.45
```

Fatalities from severe weather events is shown in the following table. 
```{r}
totalFatal <- aggregate(fatalData$FATALITIES, by=list(fatalData$EVTYPE), FUN=sum)
colnames(totalFatal) <- c("Event", "Fatalities")
totalFatal <- totalFatal[order(totalFatal$Fatalities, decreasing=TRUE),]
head(totalFatal)
```
Tornados, high heat, and flash floods are the events responsible for the greatest number of fatalities. 
```{r}
#             Event Fatalities
#141        TORNADO       5633
#26  EXCESSIVE HEAT       1903
#35     FLASH FLOOD        978
#57            HEAT        937
#97       LIGHTNING        816
#145      TSTM WIND        504
```
Injuries from severe weather events are shown here. 
```{r}
totalInjury <- aggregate(injuryData$INJURIES, by=list(injuryData$EVTYPE), FUN=sum)
colnames(totalInjury) <- c("Event", "Injuries")
totalInjury <- totalInjury[order(totalInjury$Injuries, decreasing=TRUE),]
head(totalInjury)
```
The highest number of injuries are caused by tornados, followed by floods and thunderstorm winds. 
```{r}
#             Event Injuries
#129        TORNADO    91346
#135      TSTM WIND     6957
#30           FLOOD     6789
#20  EXCESSIVE HEAT     6525
#85       LIGHTNING     5230
#47            HEAT     2100
```

